---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# forestables

Read and uniformize forest inventories data from the FIA (USA forest inventory), FFI (France forest
inventory) and IFN (Spain forest inventory).

## Installation

You can install the development version of forestables from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("emf-creaf/forestables")
```

## Inventories files

`forestables` offers download functions to get the inventory files from the official
websites from each inventory:

  - [FIA](https://www.fs.usda.gov/research/products/dataandtools/tools/fia-datamart)  
  - [FFI](https://inventaire-forestier.ign.fr/dataifn/?lang=en)  
  - IFN:
      - [IFN2](https://www.miteco.gob.es/es/biodiversidad/servicios/banco-datos-naturaleza/informacion-disponible/ifn2_descargas.html)
      - [IFN3](https://www.miteco.gob.es/es/biodiversidad/servicios/banco-datos-naturaleza/informacion-disponible/ifn3_bbdd_descargas_htm.html)
      (*accdb* are be used)
      - [IFN4](https://www.miteco.gob.es/es/biodiversidad/temas/inventarios-nacionales/inventario-forestal-nacional/cuarto_inventario.html)

This sometimes fails due to the pages being unavailable for one reason or another. We recommend
to manually download and unzip the data in the desired folder.  
Download functions are offered as helpers for programmatically approaches, but users should
be careful as data can be missing due to connection errors, pages updates...

## Example

Download FFI (France forest inventory) and extract all plots in the Loire department from 2015:

```{r example_run}
library(forestables)
library(dplyr)

## Downloading all FFI data (if not already)
ffi_path <- tempdir()
download_inventory("ffi", destination = ffi_path)

## Get the plots
ffi_filter_list <-
  show_plots_from("FFI", folder = ffi_path, departments = "42") |>
  filter(CAMPAGNE == 2015) |>
  create_filter_list()

loire_plots_2015 <- ffi_to_tibble(
  "42",
  years = 2015,
  filter_list = ffi_filter_list,
  folder = ffi_path
) |>
  clean_empty(c("tree", "shrub", "regen")) |>
  inventory_as_sf()

## Explore the plots
loire_plots_2015
```