---
title: "Selecting plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selecting plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
readRenviron("../.Renviron")
eval_code <- Sys.getenv("forestables_eval_articles") == "true"
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = eval_code
)
```

## Setup

```{r warning_message, echo = FALSE, eval = TRUE}
if (!eval_code) {
  cli::cli_inform(c(
    "i" = "Code in this vignette can't be executed without access to the data files",
    "i" = "For a complete version of this vignette, see https://emf-creaf.github.io/forestables/articles/selecting_plots.html"
  ))
}
```

```{r doc_setup}
# libraries
library(forestables)
library(dplyr)
library(sf)
library(ggplot2)
library(stringr)
library(future)
```

This vignette explains how to retrieve information about available inventory plots
and using it to select them based to use in modelling or analyses.

Workflows with all three inventories (FFI, FIA and IFN) are shown below


## FFI workflow

For the French forest inventory (FFI), we are gonna select all mediterranean plots
sampled in 2012. First step is to select de departments the plots we want are in.

```{r ffi_mediterranean_deps}
# choose the departments, those in the mediterranean area
mediterranean_deps <- c(
  "04", "06", "11", "13", "30",
  "34", "66", "83", "84"
)
```

Now we can use the department codes to obtain some basic metadata (plot codes,
coordinates and campaign year) with `show_plots_from()`.
For this we need the inventory we want plots from, the folder containing the
downloaded inventory files and a vector with the department codes (as created
above):

```{r ffi_selection}
# download the FFI data if not already (In this example the data is downloaded
# in a temporal folder and is removed after the R session ends)
ffi_folder <- tempdir()
download_inventory("FFI", ffi_folder)
# get the plots for those departments
mediterranean_plots <- show_plots_from(
  inventory = "FFI",
  folder = ffi_folder,
  departments = mediterranean_deps
)
mediterranean_plots
```

These are all the plots present in those departments for all
years, but we want 2012 plots, so we filter the results by campaign year:

```{r ffi_selection_2012}
mediterranean_plots_2012 <- mediterranean_plots |>
  filter(CAMPAGNE == 2012)
mediterranean_plots_2012
```

Now we can visualize the plots selected:

```{r ffi_plots_loc, fig.width=6, fig.align="center"}
ggplot(mediterranean_plots_2012) +
  geom_sf(
    aes(color = DEP), alpha = 0.4,
    show.legend = FALSE
  ) +
  scale_color_manual(values = hcl.colors(9, palette = "Zissou 1"))
```

Seems like we have the plots we need, so we are going to create a *filter list*
to use it to retrieve those plots data from the inventory files:

```{r ffi_filter_list_creation}
ffi_filter_list <- mediterranean_plots_2012 |>
  create_filter_list()

## we can do all in one pipe:
# ffi_filter_list <-
#   show_plots_from("FFI", folder = ffi_folder, departments = mediterranean_deps) |>
#   filter(CAMPAGNE == 2012) |>
#   create_filter_list()
```

And we are ready for retrieving the data from the plots with `ffi_to_tibble()`. For
that we need the department codes, the year, the filter list we obtained before and
the path to the folder containing the inventory files.

```{r ffi_data_2012, cache=TRUE}
## We can parallelize the process with the future package.
## In this case we use 4 parallel processes
library(future)
plan("multisession", workers = 4)

# get the data for the selected plots
ffi_data_2012 <- ffi_to_tibble(
  departments = mediterranean_deps,
  years = 2012,
  filter_list = ffi_filter_list,
  folder = ffi_folder
)

ffi_data_2012
```

Now we can explore the data, for example the most abundant species
and their mean diameter for each plot they are in:

```{r, fig.width=7, fig.height=10, fig.align="center"}
ffi_data_2012 |>
  # clean plots without tree data
  clean_empty(c("tree")) |>
  # convert inventory table to sf/spatial object
  inventory_as_sf() |>
  # unnest the tree data
  unnest("tree") |>
  # calculate dbh by species in each plot
  dplyr::group_by(id_unique_code, sp_name) |>
  dplyr::summarise(dbh = mean(dbh, na.rm = TRUE), dep = unique(dep), n = n()) |>
  dplyr::filter(!is.nan(dbh)) |>
  # filter species with 25 or less entries (plots)
  dplyr::group_by(sp_name) |>
  dplyr::mutate(n = n()) |>
  dplyr::filter(n > 15) |>
  # let's plot
  ggplot() +
  geom_sf(aes(geometry = geometry, color = dbh), size = 2.2, alpha = 0.8) +
  scale_color_gradientn(colors = hcl.colors(360, "PinkYl", rev = TRUE)) +
  facet_wrap(~sp_name, ncol = 3) +
  theme_minimal()
```

## FIA workflow

For the USA forest inventory (FIA) we are going to explore the Hawaii
inventory plots for 2019. The steps are very similar to the FFI example. First we
need the state code, retrieve the plots metadata and filter for the desired year,
2019:

```{r fia_hawaii}
# download the FIA data if not already (In this example the data is downloaded
# in a temporal folder and is removed after the R session ends)
fia_folder <- tempdir()
download_inventory("FIA", fia_folder, states = "HI")

# get the plots for those departments
hawaii_plots_2019 <- show_plots_from(
  inventory = "FIA",
  folder = fia_folder,
  states = "HI"
) |>
  filter(INVYR == 2019)
hawaii_plots_2019
```

We can visualize the plots:

```{r fia_states_polygons, fig.width=6, fig.align="center"}
# USA map for plotting
ggplot(hawaii_plots_2019) +
  geom_sf(aes(color = STATEAB, alpha = 0.4)) +
  coord_sf(xlim = c(-160.6, -154.5), ylim = c(18.5, 22.5)) +
  scale_color_manual(values = hcl.colors(1, palette = "Zissou 1"))
```

Now that we have the inventory plots selected, we can obtain the *filter list*
to retrieve the data only from those plots:

```{r fia_filter_list}
fia_filter_list <- hawaii_plots_2019 |>
  create_filter_list()

## we can do all in one pipe:
# fia_filter_list <-
#   show_plots_from("FIA", folder = fia_folder, states = "HI") |>
#   filter(INVYR == 2019) |>
#   create_filter_list()
```

And now we have everything we need for retrieving the data with
`fia_to_tibble()`. For that we need the state codes, the year, the filter list
we obtained before and the path to the folder containing the inventory files.

```{r fia_data_2019, cache=TRUE}
## We can parallelize the process with the future package.
## In this case we use 4 parallel processes
library(future)
plan("multisession", workers = 4)

# get the data
fia_data_2019 <- fia_to_tibble(
  states = "HI",
  years = 2019,
  filter_list = fia_filter_list,
  folder = fia_folder
)
```

As we did with the French inventory, we can now explore the data,
for example, the most abundant species and their mean diameter:

```{r, fig.height=8, fig.width=7, fig.align="center"}
fia_data_2019 |>
  # clean plots without tree data
  clean_empty(c("tree")) |>
  # convert inventory table to sf/spatial object
  inventory_as_sf() |>
  # unnest the tree data
  unnest("tree") |>
  # calculate dbh by species in each plot
  dplyr::group_by(id_unique_code, sp_name) |>
  dplyr::summarise(
    dbh = mean(dbh, na.rm = TRUE),
    state_ab = unique(state_ab), n = n()
  ) |>
  dplyr::filter(!is.nan(dbh)) |>
  # filter species with 100 or less entries (plots)
  dplyr::group_by(sp_name) |>
  dplyr::mutate(n = n()) |>
  dplyr::filter(n > 20) |>
  # let's plot
  ggplot() +
  geom_sf(aes(geometry = geometry, color = dbh), size = 2.2, alpha = 0.8) +
  coord_sf(xlim = c(-160.6, -154.5), ylim = c(18.5, 22.5)) +
  scale_color_gradientn(colors = hcl.colors(360, "PinkYl", rev = TRUE)) +
  facet_wrap(~sp_name, ncol = 3) +
  theme_minimal()
```

## IFN workflow

For the Spanish forest inventory (IFN) we will explore the Cantabric coast inventory
plots. The IFN, differently from other inventories, doesn't offer the data by years, but by
*versions*. We have *"ifn2"* (early 90s), *"ifn3" (early 00s)* and *"ifn4"* (2014-15).  
We are going o retrive the plots for the early 00s version (*"ifn3"*):

```{r ifn_selection}
# choose the provinces, those in the Cantabric area
north_spain_provinces <- c("24", "33", "15", "27", "39", "48", "20")

# download the IFN data if not already (In this example the data is downloaded
# in a temporal folder and is removed after the R session ends)
ifn_folder <- tempdir()
download_inventory("IFN", ifn_folder)

# get the plots for those departments
north_spain_plots <- show_plots_from(
  "IFN",
  folder = ifn_folder,
  provinces = north_spain_provinces, versions = "ifn3"
)
north_spain_plots
```

To visualize the plots:

```{r ifn_plots_visualization, fig.width=6, fig.align="center"}
ggplot(north_spain_plots) +
  geom_sf(
    aes(color = province_name_original), alpha = 0.4,
    show.legend = FALSE
  ) +
  scale_color_manual(values = hcl.colors(9, palette = "Zissou 1"))
```

Now that we have the inventory plots selected, we can obtain the *filter list*
to retrieve the data only from those plots:

```{r ifn_filter_list}
ifn_filter_list <- north_spain_plots |>
  create_filter_list()

## we can do all in one pipe:
# ifn_filter_list <- show_plots_from(
#    "IFN", folder = ifn_folder, provinces = north_spain_provinces, versions = "ifn3"
# ) |>
#   create_filter_list()
```

And now we have everything we need for retrieving the data with
`ifn_to_tibble()`:

```{r, ifn3_data, cache=TRUE}
## We can parallelize the process with the future package.
## In this case we use 4 parallel processes
library(future)
plan("multisession", workers = 4)

# get the data
ifn_data <- ifn_to_tibble(
  provinces = north_spain_provinces,
  versions = "ifn3",
  filter_list = ifn_filter_list,
  folder = ifn_folder
)
```

As we did with the French and USA inventories, we can now explore the data,
for example, the most abundant species and their mean diameter:

```{r, fig.height=7, fig.width=9, fig.align="center"}
ifn_data |>
  # clean plots without tree data
  clean_empty(c("tree")) |>
  # convert inventory table to sf/spatial object
  inventory_as_sf() |>
  # unnest the tree data
  unnest("tree") |>
  # calculate dbh by species in each plot
  dplyr::group_by(id_unique_code, sp_name) |>
  dplyr::summarise(
    dbh = mean(dbh, na.rm = TRUE),
    province_name_original = unique(province_name_original), n = n()
  ) |>
  dplyr::filter(!is.nan(dbh), !is.na(sp_name)) |>
  # filter species with 100 or less entries (plots)
  dplyr::group_by(sp_name) |>
  dplyr::mutate(n = n()) |>
  dplyr::filter(n > 500) |>
  # let's plot
  ggplot() +
  geom_sf(aes(geometry = geometry, color = dbh), size = 2.2, alpha = 0.8) +
  scale_color_gradientn(colors = hcl.colors(360, "PinkYl", rev = TRUE)) +
  facet_wrap(~sp_name, ncol = 3) +
  theme_minimal()
```
