---
title: "Selecting plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selecting plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
readRenviron("../.Renviron")
eval_code <- Sys.getenv("fia_path") != "" &&
  Sys.getenv("ffi_path") != "" &&
  Sys.getenv("ifn_path") != "" &&
  Sys.getenv("forestables_eval_articles") == "true"
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = eval_code
)
```

## Setup

```{r doc_setup, include = FALSE}
# libraries
library(forestables)
library(dplyr)
library(rnaturalearth)
library(sf)
library(ggplot2)
library(stringr)
library(future)

# paths
ffi_folder <- Sys.getenv("ffi_path")
fia_folder <- Sys.getenv("fia_path")
ifn_folder <- Sys.getenv("ifn_path")
```

This vignette explains how to retrieve information about available inventory plots
and using it to select them based to use in modelling or analyses.

Workflows with all three inventories (FFI, FIA and IFN) are shown below


## FFI workflow

For the French forest inventory (FFI), we are gonna select all mediterranean plots
sampled in 2012. First step is to select de departments the plots we want are in.

```{r ffi_mediterranean_deps}
# choose the departments, those in the mediterranean area
mediterranean_deps <- c(
  "04", "06", "11", "13", "30",
  "34", "66", "83", "84"
)
```

Now we can use the department codes to obtain some basic metadata (plot codes,
coordinates and campaign year) with `show_plots_from()`.
For this we need the inventory we want plots from, the folder containing the
downloaded inventory files and a vector with the department codes (as created
above):

```{r ffi_selection}
# get the plots for those departments
mediterranean_plots <- show_plots_from(
  inventory = "FFI",
  folder = ffi_folder,
  departments = mediterranean_deps
)
mediterranean_plots
```

These are all the plots present in those departments for all
years, but we want 2012 plots, so we filter the results by campaign year:

```{r ffi_selection_2012}
mediterranean_plots_2012 <- mediterranean_plots |>
  filter(CAMPAGNE == 2012)
mediterranean_plots_2012
```

Also, we need a the polygons for the departments for plotting results later.
To obtain them, we are gonna use the [`rnaturalearth`]() R package:

```{r ffi_department_polygons}
# france map for plotting
south_france_map <- ne_states(geounit = "france") |>
  select(dep_code = "iso_3166_2", dep_name = "gn_name") |>
  mutate(dep_code = str_remove(dep_code, "FR-")) |>
  filter(dep_code %in% mediterranean_deps)
south_france_map
```

Now we can visualize the plots selected:

```{r ffi_plots_loc, fig.width=6, fig.align="center"}
ggplot(south_france_map) +
  geom_sf() +
  geom_sf(
    aes(color = DEP), alpha = 0.4, data = mediterranean_plots_2012,
    show.legend = FALSE
  ) +
  scale_color_manual(values = hcl.colors(9, palette = "Zissou 1"))
```

Seems like we have the plots we need, so we are going to create a *filter list*
to use it to retrieve those plots data from the inventory files:

```{r ffi_filter_list_creation}
ffi_filter_list <- mediterranean_plots_2012 |>
  create_filter_list()

## we can do all in one pipe:
# ffi_filter_list <-
#   show_plots_from("FFI", folder = ffi_folder, departments = mediterranean_deps) |>
#   filter(CAMPAGNE == 2012) |>
#   create_filter_list()
```

And we are ready for retrieving the data from the plots with `ffi_to_tibble()`. For
that we need the department codes, the year, the filter list we obtained before and
the path to the folder containing the inventory files.

```{r ffi_data_2012, cache=TRUE}
## We can parallelize the process with the future package.
## In this case we use 4 parallel processes
library(future)
plan("multisession", workers = 4)

# get the data for the selected plots
ffi_data_2012 <- ffi_to_tibble(
  departments = mediterranean_deps,
  years = 2012,
  filter_list = ffi_filter_list,
  folder = ffi_folder
)

ffi_data_2012
```

Now we can explore the data, for example the most abundant species
and their mean diameter for each plot they are in:

```{r, fig.width = 6, fig.height=12, fig.align="center"}
ffi_data_2012 |>
  # clean plots without tree data
  clean_empty(c("tree")) |>
  # convert inventory table to sf/spatial object
  inventory_as_sf() |>
  # unnest the tree data
  unnest("tree") |>
  # calculate dia by species in each plot
  dplyr::group_by(id_unique_code, sp_name) |>
  dplyr::summarise(dia = mean(dia, na.rm = TRUE), dep = unique(dep), n = n()) |>
  dplyr::filter(!is.nan(dia)) |>
  # filter species with 25 or less entries (plots)
  dplyr::group_by(sp_name) |>
  dplyr::mutate(n = n()) |>
  dplyr::filter(n > 15) |>
  # let's plot
  ggplot() +
  geom_sf(data = south_france_map, alpha = 0, linewidth = 0.3) +
  geom_sf(aes(geometry = geometry, color = dia), size = 2.2, alpha = 0.8) +
  scale_color_gradientn(colors = hcl.colors(360, "PinkYl", rev = TRUE)) +
  facet_wrap(~sp_name, ncol = 3) +
  theme_minimal()
```

## FIA workflow

For the USA forest inventory (FIA) we are going to explore the west coast
inventory plot for 2012. The steps are very similar to the FFI example. First we
need the state codes, retrieve the plots metadata for those states and filter for
the desired year, 2012::

```{r fia_west_coast_states}
# choose the provinces, those in the Cantabric area
west_coast_states <- c("WA", "CA", "OR")
# get the plots for those departments
west_coast_plots_2012 <- show_plots_from(
  inventory = "FIA",
  folder = fia_folder,
  states = west_coast_states
) |>
  filter(INVYR == 2012)
west_coast_plots_2012
```

Also, if we obtain the states polygons with the `rnaturalearth` library, we
can visualize the plots:

```{r fia_states_polygons, fig.width=6, fig.align="center"}
# USA map for plotting
west_coast_map <- ne_states(geounit = "United States of America") |>
  select(state_code = "iso_3166_2", state_name = "gn_name") |>
  mutate(state_code = str_remove(state_code, "US-")) |>
  filter(state_code %in% west_coast_states)

ggplot(west_coast_map) +
  geom_sf() +
  geom_sf(aes(color = STATEAB, alpha = 0.4), data = west_coast_plots_2012) +
  scale_color_manual(values = hcl.colors(3, palette = "Zissou 1"))
```

Now that we have the inventory plots selected, we can obtain the *filter list*
to retrieve the data only from those plots:

```{r fia_filter_list}
fia_filter_list <- west_coast_plots_2012 |>
  create_filter_list()

## we can do all in one pipe:
# fia_filter_list <-
#   show_plots_from("FIA", folder = fia_folder, states = west_coast_states) |>
#   filter(INVYR == 2012) |>
#   create_filter_list()
```

And now we have everything we need for retrieving the data with
`fia_to_tibble()`. For that we need the state codes, the year, the filter list
we obtained before and the path to the folder containing the inventory files.

```{r fia_data_2012, cache=TRUE}
## We can parallelize the process with the future package.
## In this case we use 4 parallel processes
library(future)
plan("multisession", workers = 4)

# get the data
fia_data_2012 <- fia_to_tibble(
  states = west_coast_states,
  years = 2012,
  filter_list = fia_filter_list,
  folder = fia_folder
)
```

As we did with the French inventory, we can now explore the data,
for example, the most abundant species and their mean diameter:

```{r, fig.height=12, fig.width=6, fig.align="center"}
fia_data_2012 |>
  # clean plots without tree data
  clean_empty(c("tree")) |>
  # convert inventory table to sf/spatial object
  inventory_as_sf() |>
  # unnest the tree data
  unnest("tree") |>
  # calculate dia by species in each plot
  dplyr::group_by(id_unique_code, sp_name) |>
  dplyr::summarise(
    dia = mean(dia, na.rm = TRUE),
    state_ab = unique(state_ab), n = n()
  ) |>
  dplyr::filter(!is.nan(dia)) |>
  # filter species with 100 or less entries (plots)
  dplyr::group_by(sp_name) |>
  dplyr::mutate(n = n()) |>
  dplyr::filter(n > 100) |>
  # let's plot
  ggplot() +
  geom_sf(data = west_coast_map, alpha = 0, linewidth = 0.3) +
  geom_sf(aes(geometry = geometry, color = dia), size = 2.2, alpha = 0.8) +
  scale_color_gradientn(colors = hcl.colors(360, "PinkYl", rev = TRUE)) +
  facet_wrap(~sp_name, ncol = 6) +
  theme_minimal()
```

## IFN workflow

For the Spanish forest inventory (IFN) we will explore the Cantabric coast inventory
plots. The IFN, differently from other inventories, doesn't offer the data by years, but by
*versions*. We have *"ifn2"* (early 90s), *"ifn3" (early 00s)* and *"ifn4"* (2014-15).  
We are going o retrive the plots for the latest version (*"ifn4"*):

```{r ifn_selection}
# choose the provinces, those in the Cantabric area
north_spain_provinces <- c("24", "33", "15", "27", "39", "48", "20")
# get the plots for those departments
north_spain_plots <- show_plots_from(
  "IFN",
  folder = ifn_folder,
  provinces = north_spain_provinces, versions = "ifn4"
)
north_spain_plots
```

Also, we obtain the provinces polygins with `rnaturalearth`, to visualize the
plots:

```{r ifn_plots_visualization, fig.width=6, fig.align="center"}
# spain map for plotting
north_spain_map <- ne_states(country = "spain") |>
  select(dep_code = "iso_3166_2", dep_name = "name") |>
  mutate(dep_code = str_remove(dep_code, "ES-")) |>
  filter(dep_name %in% c(
    "León", "Asturias", "La Coruña", "Lugo", "Cantabria", "Bizkaia", "Gipuzkoa"
  ))

ggplot(north_spain_map) +
  geom_sf() +
  geom_sf(
    aes(color = province_name_original), data = north_spain_plots, alpha = 0.4,
    show.legend = FALSE
  ) +
  scale_color_manual(values = hcl.colors(9, palette = "Zissou 1"))
```

Now that we have the inventory plots selected, we can obtain the *filter list*
to retrieve the data only from those plots:

```{r ifn_filter_list}
ifn_filter_list <- north_spain_plots |>
  create_filter_list()

## we can do all in one pipe:
# ifn_filter_list <- show_plots_from(
#    "IFN", folder = ifn_folder, provinces = north_spain_provinces, versions = "ifn4"
# ) |>
#   create_filter_list()
```

And now we have everything we need for retrieving the data with
`ifn_to_tibble()`:

```{r, ifn4_data, cache=TRUE}
## We can parallelize the process with the future package.
## In this case we use 4 parallel processes
library(future)
plan("multisession", workers = 4)

# get the data
ifn_data <- ifn_to_tibble(
  provinces = north_spain_provinces,
  versions = "ifn4",
  filter_list = ifn_filter_list,
  folder = ifn_folder
)
```

As we did with the French and USA inventories, we can now explore the data,
for example, the most abundant species and their mean diameter:

```{r, fig.height=12, fig.width=6, fig.align="center"}
ifn_data |>
  # clean plots without tree data
  clean_empty(c("tree")) |>
  # convert inventory table to sf/spatial object
  inventory_as_sf() |>
  # unnest the tree data
  unnest("tree") |>
  # calculate dia by species in each plot
  dplyr::group_by(id_unique_code, sp_name) |>
  dplyr::summarise(
    dia = mean(dia, na.rm = TRUE),
    province_name_original = unique(province_name_original), n = n()
  ) |>
  dplyr::filter(!is.nan(dia)) |>
  # filter species with 100 or less entries (plots)
  dplyr::group_by(sp_name) |>
  dplyr::mutate(n = n()) |>
  dplyr::filter(n > 500) |>
  # let's plot
  ggplot() +
  geom_sf(data = north_spain_map, alpha = 0, linewidth = 0.3) +
  geom_sf(aes(geometry = geometry, color = dia), size = 2.2, alpha = 0.8) +
  scale_color_gradientn(colors = hcl.colors(360, "PinkYl", rev = TRUE)) +
  facet_wrap(~sp_name, ncol = 4) +
  theme_minimal()
```
