---
title: "Selecting plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selecting plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

```{r setup}
# libraries
library(esus)
library(sf)
library(dplyr)
library(rnaturalearth)
library(ggplot2)
library(stringr)
library(furrr)
library(future)

# paths
ffi_folder <- Sys.getenv("ffi_path")
fia_folder <- Sys.getenv("fia_path")
ifn_folder <- Sys.getenv("ifn_path")
```


## FFI example

```{r ffi_selection}
# choose the departments, those in the mediterranean area
mediterranean_deps <- c("66", "11", "34", "30", "13", "84", "04", "83", "06")
# get the plots for those departments
mediterranean_plots <- show_plots_from("FFI", folder = ffi_folder, departments = mediterranean_deps) 
mediterranean_plots
```


```{r plots_visualization}
# france map for plotting
france_map <- ne_states(geounit = "france") |>
  select(dep_code = "iso_3166_2", dep_name = "gn_name") |>
  mutate(dep_code = str_remove(dep_code, "FR-")) |>
  filter(dep_code %in% mediterranean_deps)

ggplot(france_map) +
  geom_sf() +
  geom_sf(aes(color = DEP, alpha = 0.1), data = mediterranean_plots) +
  scale_color_manual(values = hcl.colors(9, palette = "Zissou 1"))

mediterranean_plots_2012 <- mediterranean_plots |>
  filter(CAMPAGNE == 2012)

ggplot(france_map) +
  geom_sf() +
  geom_sf(aes(color = DEP, alpha = 0.1), data = mediterranean_plots_2012) +
  scale_color_manual(values = hcl.colors(9, palette = "Zissou 1"))

filter_list_2012 <-
  show_plots_from("FFI", folder = ffi_folder, departments = mediterranean_deps) |>
  filter(CAMPAGNE == 2012) |>
  create_filter_list()


plan("multisession")
data_2012 <- ffi_to_tibble(
  departments = mediterranean_deps,
  years = 2012,
  filter_list = filter_list_2012,
  folder = ffi_folder
)

data_2012 |>
  dplyr::filter(!purrr::map_lgl(data_2012$tree, rlang::is_empty)) |>
  sf::st_as_sf(coords = c("COORD1", "COORD2"), crs = unique(data_2012$crs)) |>
  sf::st_transform(crs = 4326) |>
  tidyr::unnest("tree") |>
  dplyr::group_by(ID_UNIQUE_PLOT, SP_NAME) |>
  dplyr::summarise(DENSITY = mean(DENSITY, na.rm = TRUE), DEP = unique(DEP), n = n()) |>
  dplyr::filter(!is.nan(DENSITY)) |>
  dplyr::group_by(SP_NAME) |>
  dplyr::mutate(n = n()) |>
  dplyr::filter(n > 25, DENSITY < 300) |>
  dplyr::left_join(as_tibble(france_map), by = c("DEP" = "dep_code")) |>
  ggplot() +
  geom_sf(aes(geometry = geometry.y), data = ~ distinct(as_tibble(.x)[,c("geometry.y")]), alpha = 0, linewidth = 0.3) +
  geom_sf(aes(geometry = geometry.x, color = DENSITY), size = 2.2, alpha = 0.8) +
  scale_color_gradientn(colors = hcl.colors(360, "PinkYl", rev = TRUE)) +
  facet_wrap(~ SP_NAME) +
  theme_minimal()
```


### FFI Tree diagnostics

```{r tree_diagnostics}
# understory
data_2012 |>
  dplyr::filter(!purrr::map_lgl(data_2012$tree, rlang::is_empty)) |>
  sf::st_as_sf(coords = c("COORD1", "COORD2"), crs = unique(data_2012$crs)) |>
  sf::st_transform(crs= 4326) |>
  tidyr::unnest(understory, names_sep = "_") |>
  tidyr::unnest(understory_herbs, names_sep = "_") |>
  ggplot() +
  geom_sf(data = france_map) +
  geom_sf(aes(color = understory_herbs_HERB)) +
  scale_color_gradientn(colors = hcl.colors(20, "SunsetDark"))


data_clean_2012 <- data_2012 |>
  dplyr::filter(!purrr::map_lgl(data_2012$tree, rlang::is_empty)) |>
  sf::st_as_sf(coords = c("COORD1", "COORD2"), crs = unique(data_2012$crs)) |>
  sf::st_transform(crs= 4326)

na_in_diameter_index <- purrr::map_lgl(
  seq_len(nrow(data_clean_2012)),
  .f = \(i) {
    data_clean_2012 |>
      dplyr::pull(tree) |>
      magrittr::extract2(i) |>
      dplyr::pull(DIA) |>
      is.na() |>
      all()
  }
)

na_in_density_index <- purrr::map_lgl(
  seq_len(nrow(data_clean_2012)),
  .f = \(i) {
    data_clean_2012 |>
      dplyr::pull(tree) |>
      magrittr::extract2(i) |>
      dplyr::pull(DENSITY) |>
      is.na() |>
      all()
  }
)

na_in_ht_index <- purrr::map_lgl(
  seq_len(nrow(data_clean_2012)),
  .f = \(i) {
    data_clean_2012 |>
      dplyr::pull(tree) |>
      magrittr::extract2(i) |>
      dplyr::pull(HT) |>
      is.na() |>
      all()
  }
)

na_in_sp_name_index <-  purrr::map_lgl(
  seq_len(nrow(data_clean_2012)),
  .f = \(i) {
    data_clean_2012 |>
      dplyr::pull(tree) |>
      magrittr::extract2(i) |>
      dplyr::pull(SP_NAME) |>
      is.na() |>
      all()
  }
)

identical(na_in_diameter_index, na_in_ht_index)
sum(na_in_diameter_index); sum(na_in_ht_index)
sum(na_in_sp_name_index)


data_clean_2012 |>
  tidyr::unnest(tree, names_sep = "_")
```

### FFI benchmarking

From previous benchmarks, grep still is the better option, even if there is no cache for files
read.

This bechmark finds the better configuration of paralellization for the example data set

```{r ffi_bench, eval=FALSE}
plan("multisession")
ffi_to_tibble_benchmark <- bench::mark(
  sche_2 = ffi_to_tibble(
    departments = mediterranean_deps,
    years = 2012,
    filter_list = filter_list_2012,
    folder = ffi_folder,
    .parallel_options = furrr_options(scheduling = 1L), .verbose = TRUE
  ),
  sche_1 = ffi_to_tibble(
    departments = mediterranean_deps,
    years = 2012,
    filter_list = filter_list_2012,
    folder = ffi_folder,
    .parallel_options = furrr_options(scheduling = 1L), .verbose = TRUE
  ),
  chunk_300 = ffi_to_tibble(
    departments = mediterranean_deps,
    years = 2012,
    filter_list = filter_list_2012,
    folder = ffi_folder,
    .parallel_options = furrr_options(chunk_size = 300L), .verbose = TRUE
  ),
  chunk_100 = ffi_to_tibble(
    departments = mediterranean_deps,
    years = 2012,
    filter_list = filter_list_2012,
    folder = ffi_folder,
    .parallel_options = furrr_options(chunk_size = 100L), .verbose = TRUE
  ),
  min_time = Inf, iterations = 3
)
saveRDS(
  ffi_to_tibble_benchmark,
  file = paste0("data-raw/benchmarking/ffi_to_tibble_", Sys.Date(), ".rds")
)

ffi_to_tibble_benchmark
```

Seems like the best is around chunk 100, but the standard, scheduling 1 is also good. No big
differences between parallelization strategies.


## FIA

```{r fia_selection}
# choose the departments, those in the mediterranean area
west_coast_states <- c("WA", "CA", "OR")
# get the plots for those departments
west_coast_plots <- show_plots_from("FIA", folder = fia_folder, states = west_coast_states) 
west_coast_plots
```


```{r fia_plots_visualization}
# france map for plotting
west_coast_map <- ne_states(geounit = "United States of America") |>
  select(dep_code = "iso_3166_2", dep_name = "gn_name") |>
  mutate(dep_code = str_remove(dep_code, "US-")) |>
  filter(dep_code %in% west_coast_states)

ggplot(west_coast_map) +
  geom_sf() +
  geom_sf(aes(color = STATEAB, alpha = 0.1), data = west_coast_plots) +
  scale_color_manual(values = hcl.colors(9, palette = "Zissou 1"))

west_coast_plots_2012 <- west_coast_plots |>
  filter(INVYR == 2012)

ggplot(west_coast_map) +
  geom_sf() +
  geom_sf(aes(color = STATEAB, alpha = 0.1), data = west_coast_plots_2012) +
  scale_color_manual(values = hcl.colors(9, palette = "Zissou 1"))

fia_filter_list_2012 <-
  show_plots_from("FIA", folder = fia_folder, states = west_coast_states) |>
  filter(INVYR == 2012) |>
  create_filter_list()

# fia_filter_list_2012 <- fia_filter_list_2012 |>
#     purrr::map_depth(.depth = 2, .f = \(x) {
#       sample(x, 5, replace = TRUE) |> unique()
#     })

plan("multisession")
fia_data_2012 <- fia_to_tibble(
  states = west_coast_states,
  years = 2012,
  filter_list = fia_filter_list_2012,
  folder = fia_folder
)

fia_data_2012 |>
  dplyr::filter(!purrr::map_lgl(fia_data_2012$tree, rlang::is_empty)) |>
  sf::st_as_sf(coords = c("COORD1", "COORD2"), crs = unique(fia_data_2012$COORD_SYS)) |>
  sf::st_transform(crs = 4326) |>
  as_tibble() |>
  tidyr::unnest("tree") |>
  dplyr::group_by(ID_UNIQUE_PLOT, SP_NAME) |>
  dplyr::summarise(DENSITY = mean(DENSITY, na.rm = TRUE), STATEAB = unique(STATEAB), n = n(), geometry = first(geometry)) |>
  dplyr::filter(!is.nan(DENSITY)) |>
  dplyr::group_by(SP_NAME) |>
  dplyr::mutate(n = n()) |>
  dplyr::filter(n > 200, DENSITY < 300) |>
  dplyr::left_join(as_tibble(west_coast_map), by = c("STATEAB" = "dep_code")) |>
  ggplot() +
  geom_sf(aes(geometry = geometry.y), data = ~ distinct(as_tibble(.x)[,c("geometry.y")]), alpha = 0, linewidth = 0.3) +
  geom_sf(aes(geometry = geometry.x, color = DENSITY), size = 2.2, alpha = 0.8) +
  scale_color_gradientn(colors = hcl.colors(360, "PinkYl", rev = TRUE)) +
  facet_wrap(~ SP_NAME) +
  theme_minimal()

fia_data_2012 |>
  dplyr::filter(!purrr::map_lgl(fia_data_2012$regen, rlang::is_empty)) |>
  sf::st_as_sf(coords = c("COORD1", "COORD2"), crs = unique(fia_data_2012$COORD_SYS)) |>
  sf::st_transform(crs = 4326) |>
  as_tibble() |>
  tidyr::unnest("regen") |>
  dplyr::group_by(ID_UNIQUE_PLOT, SP_NAME) |>
  dplyr::summarise(DENSITY = mean(DENSITY, na.rm = TRUE), STATEAB = unique(STATEAB), n = n(), geometry = first(geometry)) |>
  dplyr::filter(!is.nan(DENSITY)) |>
  dplyr::group_by(SP_NAME) |>
  dplyr::mutate(n = n()) |>
  dplyr::filter(n > 100, DENSITY < 300) |>
  dplyr::left_join(as_tibble(west_coast_map), by = c("STATEAB" = "dep_code")) |>
  ggplot() +
  geom_sf(aes(geometry = geometry.y), data = ~ distinct(as_tibble(.x)[,c("geometry.y")]), alpha = 0, linewidth = 0.3) +
  geom_sf(aes(geometry = geometry.x, color = DENSITY), size = 2.2, alpha = 0.8) +
  scale_color_gradientn(colors = hcl.colors(360, "PinkYl", rev = TRUE)) +
  facet_wrap(~ SP_NAME) +
  theme_minimal()
```


## IFN

```{r ifn_selection}
# choose the departments, those in the mediterranean area
north_spain_provinces <- c("24", "33", "15", "27", "39", "48", "20")
# get the plots for those departments
north_spain_plots <- show_plots_from(
  "IFN", folder = ifn_folder,
  provinces = north_spain_provinces, versions = "ifn4"
)
north_spain_plots
```


```{r ifn_plots_visualization}
# france map for plotting
north_spain_map <- ne_states(country = "spain") |>
  select(dep_code = "iso_3166_2", dep_name = "name") |>
  mutate(dep_code = str_remove(dep_code, "ES-")) |>
  filter(dep_name %in% c(
    "León", "Asturias", "La Coruña", "Lugo", "Cantabria", "Bizkaia", "Gipuzkoa"
  ))

ggplot(north_spain_map) +
  geom_sf() +
  geom_sf(aes(color = province_name_original), data = north_spain_plots, alpha = 0.1) +
  scale_color_manual(values = hcl.colors(9, palette = "Zissou 1"))

ifn_filter_list <-
  show_plots_from(
    "IFN", folder = ifn_folder,
    provinces = north_spain_provinces, versions = "ifn4"
  ) |>
  create_filter_list()

# fia_filter_list_2012 <- fia_filter_list_2012 |>
#     purrr::map_depth(.depth = 2, .f = \(x) {
#       sample(x, 5, replace = TRUE) |> unique()
#     })

plan("multisession")
ifn_data <- ifn_to_tibble(
  provinces = north_spain_provinces,
  versions = "ifn4",
  filter_list = ifn_filter_list,
  folder = ifn_folder
)

ifn_data |>
  dplyr::filter(!purrr::map_lgl(ifn_data$tree, rlang::is_empty)) |>
  dplyr::group_by(crs) |>
  dplyr::group_modify(
    .f = \(crs_subset, crs) {
      sf::st_as_sf(crs_subset, coords = c("COORD1", "COORD2"), crs = unique(crs$crs)) |>
        sf::st_transform(crs = 4326) |>
        dplyr::mutate(crs_orig = crs$crs)
    }
  ) |>
  # dplyr::bind_rows() |>
  # magrittr::extract2(2) |>
  # as_tibble() |>
  tidyr::unnest("tree") |>
  dplyr::group_by(ID_UNIQUE_PLOT, SP_NAME) |>
  dplyr::summarise(
    DENSITY = mean(DENSITY, na.rm = TRUE),
    province_name_original = unique(province_name_original),
    n = n(), geometry = first(geometry)
  ) |>
  dplyr::filter(!is.nan(DENSITY)) |>
  dplyr::group_by(SP_NAME) |>
  dplyr::mutate(n = n()) |>
  dplyr::filter(n > 150, DENSITY < 300) |>
  # dplyr::left_join(as_tibble(north_spain_map), by = c("STATEAB" = "dep_code")) |>
  ggplot() +
  geom_sf(data = north_spain_map, alpha = 0, linewidth = 0.3) +
  geom_sf(aes(geometry = geometry, color = DENSITY), size = 2.2, alpha = 0.8) +
  scale_color_gradientn(colors = hcl.colors(360, "PinkYl", rev = TRUE)) +
  facet_wrap(~ SP_NAME) +
  theme_minimal()

ifn_data |>
  dplyr::filter(!purrr::map_lgl(ifn_data$regen, rlang::is_empty)) |>
  sf::st_as_sf(coords = c("COORD1", "COORD2"), crs = unique(ifn_data$COORD_SYS)) |>
  sf::st_transform(crs = 4326) |>
  as_tibble() |>
  tidyr::unnest("regen") |>
  dplyr::group_by(ID_UNIQUE_PLOT, SP_NAME) |>
  dplyr::summarise(DENSITY = mean(DENSITY, na.rm = TRUE), STATEAB = unique(STATEAB), n = n(), geometry = first(geometry)) |>
  dplyr::filter(!is.nan(DENSITY)) |>
  dplyr::group_by(SP_NAME) |>
  dplyr::mutate(n = n()) |>
  dplyr::filter(n > 100, DENSITY < 300) |>
  dplyr::left_join(as_tibble(west_coast_map), by = c("STATEAB" = "dep_code")) |>
  ggplot() +
  geom_sf(aes(geometry = geometry.y), data = ~ distinct(as_tibble(.x)[,c("geometry.y")]), alpha = 0, linewidth = 0.3) +
  geom_sf(aes(geometry = geometry.x, color = DENSITY), size = 2.2, alpha = 0.8) +
  scale_color_gradientn(colors = hcl.colors(360, "PinkYl", rev = TRUE)) +
  facet_wrap(~ SP_NAME) +
  theme_minimal()
```
